\thetitle

### 2025-05-23 workshop/PT.tex
ns edits

Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
in \cite{syed2021nrpt}, which we outline in \cref{alg:distributed_PT}.
This section gives both a brief overview of PT and some details of 
its distributed implementation.

>>>>

Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
in \citet{syed2021nrpt}, which we outline in \cref{alg:distributed_PT}.
This section gives both a brief overview of PT and some details of 
its distributed implementation.

### 2025-05-23 workshop/SPI.tex
compress
ns edits

\subsection{Distributed reduction and floating point non-associativity}
After each round of PT, the machines need to exchange information such as the 
average swap acceptance probabilities, statistics to adapt a variational 
reference and adjust annealing parameters, and so on. 
For instance, suppose our \texttt{state} is univariate and real-valued and 
that each \texttt{Replica} keeps track of the number of times the target chain $N$ 
is visited as well as the mean of the univariate \texttt{state}s from the target chain.
To obtain the final estimate of the mean of the target distribution, we would like 
to pool the mean estimates from each of the \texttt{Replica}s, weighted by 
the number of times that each \texttt{Replica} visited the target chain.   
This process involves summing floating-point values that are located on each 
of the \texttt{Replica}s/machines and is an example of a reduction scheme.

@@

const n_iters = 10000
result = zeros(n_iters)
Random.seed!(1)
@threads for i in 1:n_iters
    result[i] = rand()
end
println("Result: $(last(result))")
\end{lstlisting}
With eight threads, this outputs:
\begin{lstlisting}[language = Julia]
Num. of threads: 8
Result: 0.25679999169092793
\end{lstlisting}
Julia guarantees that if we rerun this code, as long as we are using eight threads, 
we will always get the same result, irrespective of the multi-threading 
scheduling decisions implied by the \texttt{@threads}-loop. 
Internally, Julia works with task-local RNGs and the \texttt{@threads} macro 
spawns \texttt{nthreads()} number of task-local RNGs.
For this reason, with a different number of threads, the result is different:
\begin{lstlisting}[language = Julia]
Num. of threads: 1
Result: 0.8785201210435906
\end{lstlisting}

@@

const n_iters = 10000
const master_rng = SplittableRandom(1)
result = zeros(n_iters)
rngs = [split(master_rng) for _ in 1:n_iters]
Random.seed!(1)
@threads for i in 1:n_iters
    result[i] = rand(rngs[i])
end
println("Result: $(last(result))")
\end{lstlisting}
With one and eight threads, the code above outputs
\begin{lstlisting}[language = Julia]
Num. of threads: 1
Result: 0.4394633333251359
\end{lstlisting}
\begin{lstlisting}[language = Julia]
Num. of threads: 8
Result: 0.4394633333251359
\end{lstlisting}

>>>>

\subsection{Distributed reduction and floating point non-associativity}
After each round of PT, 
\NS{I think we should assume that the readers don't know much about PT, so maybe 
this part should be simplified}
the machines need to exchange information such as the 
average swap acceptance probabilities, statistics to adapt a variational 
reference and adjust annealing parameters, and so on. 
For instance, suppose our \texttt{state} is univariate and real-valued and 
that each \texttt{Replica} keeps track of the number of times the target chain $N$ 
is visited as well as the mean of the univariate \texttt{state}s from the target chain.
To obtain the final estimate of the mean of the target distribution, we would like 
to pool the mean estimates from each of the \texttt{Replica}s, weighted by 
the number of times that each \texttt{Replica} visited the target chain.   
This process involves summing floating-point values that are located on each 
of the \texttt{Replica}s/machines and is an example of a reduction scheme.

@@

const n_iters = 10000
result = zeros(n_iters)
Random.seed!(1)
@threads for i in 1:n_iters
    result[i] = rand()
end
println("Result: $(last(result))")
\end{lstlisting}
With eight threads, this outputs:
\begin{lstlisting}[language = Julia]
Num. of threads: 8
Result: 0.25679999169092793
\end{lstlisting}
Julia guarantees that if we rerun this code, as long as we are using eight threads, 
we will always get the same result, irrespective of the multi-threading 
scheduling decisions implied by the \texttt{@threads}-loop. 
Internally, Julia works with task-local RNGs and the \texttt{@threads} macro 
spawns \texttt{nthreads()} number of task-local RNGs.
For this reason, with a different number of threads, the result is different:
% NB: use minipage to avoid splitting the code across pages
\begin{minipage}{\linewidth}
\begin{lstlisting}[language = Julia]
Num. of threads: 1
Result: 0.8785201210435906
\end{lstlisting}
\end{minipage}

@@

const n_iters = 10000
const master_rng = SplittableRandom(1)
result = zeros(n_iters)
rngs = [split(master_rng) for _ in 1:n_iters]
Random.seed!(1)
@threads for i in 1:n_iters
    result[i] = rand(rngs[i])
end
println("Result: $(last(result))")
\end{lstlisting}
With one and eight threads, the code above outputs
% NB: use minipage to avoid splitting the code across pages
\begin{minipage}{\linewidth}
\begin{lstlisting}[language = Julia]
Num. of threads: 1
Result: 0.4394633333251359

Num. of threads: 8
Result: 0.4394633333251359
\end{lstlisting}
\end{minipage}

### 2025-05-23 workshop/abstract.tex
TPG section + reorder
new devels + fixes
ns edits

\begin{abstract}
Sampling from complicated probability distributions---such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics---is a challenging problem. Pigeons.jl provides 
a simple user interface to perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines. In this paper,
we discuss a feature of Pigeons.jl that we call \emph{strong parallelism invariance} (SPI): 
a guarantee that the output for a given seed is identical irrespective of the number 
of threads and processes. This feature is crucial for scientific reproducibility and 
software validation. We describe the key features of Pigeons.jl that enable a 
distributed and randomized algorithm satisfying SPI.
\end{abstract}

>>>>

\begin{abstract}
\NS{Feel free to place yourself as first author, Miguel.}
Sampling from complicated probability distributions---such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics---is a challenging problem. Pigeons.jl provides 
a simple user interface to perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines. In this paper,
we discuss a feature of Pigeons.jl that we call \emph{strong parallelism invariance} (SPI): 
a guarantee that the output for a given seed is identical irrespective of the number 
of threads and processes. This feature is crucial for scientific reproducibility and 
software validation. We describe the key features of Pigeons.jl that enable a 
distributed and randomized algorithm satisfying SPI. Finally, we briefly discuss 
some recent developments in Pigeons.jl.
\end{abstract}

### 2025-05-23 workshop/examples.tex
ns edits

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering 
\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

@@

    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
    which measures the inherent difficulty of the sampling problem.
    A rule of thumb to configure the number of PT chains is also 
    given by \cite{syed2021nrpt}, where they suggest that stable performance 
    should be achieved when the number of chains is set to roughly 2$\Lambda$.
    See \cref{sec:PT_diagnostics} for more information.

@@

\subsubsection{Online statistics}
\label{sec:online_stats}
Pigeons has facilities to support cases requiring large memory. For instance, we
allow for the computation of online statistics, as well as off-memory sample storage.
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation. For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

@@

\subsubsection{PT diagnostics}
\label{sec:PT_diagnostics}
We describe how to produce some key parallel tempering diagnostics from 
\cite{syed2021nrpt}.

The global communication barrier, denoted $\Lambda$ in Pigeons.jl output, can be 
used to approximately inform the appropriate number of chains. 
Based on \cite{syed2021nrpt}, stable PT performance should be achieved when  
the number of chains is set to roughly $2\Lambda$. This can be achieved by 
modifying the \texttt{n\_chains} argument in the call to \texttt{pigeons()}. 
The global communication barrier is shown at each round and can also be accessed 
with 
\begin{lstlisting}[language=Julia]
Pigeons.global_barrier(pt)
\end{lstlisting}

@@

\subsection{Additional options}
\label{sec:additional_options}
In the preceding example we only specified the target distribution and let 
Pigeons.jl decide on default values for most other settings of the inference engine. 
There are various settings we can change, including: 
the random seed (\texttt{seed}), the number of PT chains (\texttt{n\_chains}), 
the number of PT tuning rounds/samples (\texttt{n\_rounds}), 
and a variational reference distribution family (\texttt{variational}), among other settings.
For instance, we can run 
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model),
    n_rounds = 10, 
    n_chains = 10,
    variational = GaussianReference(),
    seed = 2
)
\end{lstlisting}
which runs PT with the same Turing model target as before and explicitly states 
that we should use 10 PT tuning rounds with 10 chains (described below). 
In the above code chunk we also specify that we would like to use a 
Gaussian variational reference distribution.
That is, the reference distribution is chosen from a multivariate Gaussian family 
that lies as close as possible to the target distribution in order to improve 
the efficiency of PT. We refer readers to \cite{surjanovic2022vpt} for more details.
When only continuous parameters are of interest, we encourage users to consider 
using \texttt{variational = GaussianReference()} and setting \texttt{n\_chains\_variational = 10}, 
for example, as the number of restarts may substantially increase with these settings.

>>>>

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering 
\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \citet{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

@@

    \item \texttt{$\Lambda$}: the global communication barrier, as described in \citet{syed2021nrpt}, 
    which measures the inherent difficulty of the sampling problem.
    A rule of thumb to configure the number of PT chains is also 
    given by \citet{syed2021nrpt}, where they suggest that stable performance 
    should be achieved when the number of chains is set to roughly 2$\Lambda$.
    See \cref{sec:PT_diagnostics} for more information.

@@

\subsubsection{Online statistics}
\label{sec:online_stats}
Pigeons has facilities to support cases requiring large memory. For instance, we
allow for the computation of online statistics, as well as off-memory sample storage.
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation. 
\NS{We should link to the online documentation here and also earlier.}
For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

@@

\subsubsection{PT diagnostics}
\label{sec:PT_diagnostics}
We describe how to produce some key parallel tempering diagnostics from 
\citet{syed2021nrpt}.

The global communication barrier, denoted $\Lambda$ in Pigeons.jl output, can be 
used to approximately inform the appropriate number of chains. 
Based on \citet{syed2021nrpt}, stable PT performance should be achieved when  
the number of chains is set to roughly $2\Lambda$. This can be achieved by 
modifying the \texttt{n\_chains} argument in the call to \texttt{pigeons()}. 
The global communication barrier is shown at each round and can also be accessed 
with 
\begin{lstlisting}[language=Julia]
Pigeons.global_barrier(pt)
\end{lstlisting}

@@

\subsection{Additional options}
\label{sec:additional_options}
In the preceding example we only specified the target distribution and let 
Pigeons.jl decide on default values for most other settings of the inference engine. 
There are various settings we can change, including: 
the random seed (\texttt{seed}), the number of PT chains (\texttt{n\_chains}), 
the number of PT tuning rounds/samples (\texttt{n\_rounds}), 
and a variational reference distribution family (\texttt{variational}), among other settings.
For instance, we can run 
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model),
    n_rounds = 10, 
    n_chains = 10,
    variational = GaussianReference(),
    seed = 2
)
\end{lstlisting}
which runs PT with the same Turing model target as before and explicitly states 
that we should use 10 PT tuning rounds with 10 chains (described below). 
In the above code chunk we also specify that we would like to use a 
Gaussian variational reference distribution.
That is, the reference distribution is chosen from a multivariate Gaussian family 
that lies as close as possible to the target distribution in order to improve 
the efficiency of PT. We refer readers to \citet{surjanovic2022vpt} for more details.
When only continuous parameters are of interest, we encourage users to consider 
using \texttt{variational = GaussianReference()} and setting \texttt{n\_chains\_variational = 10}, 
for example, as the number of restarts may substantially increase with these settings.

### 2025-05-23 workshop/introduction.tex
ns edits

Pigeons.jl\footnote{\url{https://github.com/Julia-Tempering/Pigeons.jl}}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible parallel tempering
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

>>>>

Pigeons.jl\footnote{\url{https://github.com/Julia-Tempering/Pigeons.jl}}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible parallel tempering
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1,000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}
